# Text Pre-Processing: String manipulation

```{r echo=FALSE}
vembedr::embed_youtube("GShnUE8r4Io")
```


When working with data, a significant number of variables will be in some sort of text format. When you want to manipulate these variables, an easy approach would be exporting the data to MS Excel and then just performing those manipulations by hand. This is very time-consuming, though, and, hence, I rather recommend the R way which scales well and works fast for data sets of varying sizes.

Quick reminder: a string is an element of a character vector and can be created by simply wrapping some text in quotation marks:

```{r}
string <- "Hi, how are you doing?"
vector_of_strings <- c("Hi, how are you doing?", "I'm doing well, HBY?", "Me too, thanks for asking.")
```

Note that you can either wrap your text in double quotation marks and use single ones in the string and vice versa:

```{r}
single_ones <- "what's up"
double_ones <- 'he said: "I am fine"'
```

The `stringr` package [@wickham2019b] contains a multitude of commands (49 in total) that can be used to achieve a couple of things, mainly manipulating character vectors, and finding and matching patterns. Basically, these goals can also be achieved with base R functions, but `stringr`'s advantage is its consistency. The makers of `stringr` describe it as

> A consistent, simple and easy to use set of wrappers around the fantastic `stringi` package. All function and argument names (and positions) are consistent, all functions deal with `NA`'s and zero-length vectors in the same way, and the output from one function is easy to feed into the input of another.

Every `stringr` function starts with `str_` -- which facilitates finding the proper command: just type `str_` and RStudio's auto-suggest function should take care of the rest (if it doesn't pop up by itself, you can trigger it by hitting the tab key). Also, they take a vector of strings as their first argument, which facilitates using them in a ` %>% `-pipeline and adding them to a `mutate()`-call. 

One important component of `stringr` functions is regular expressions which will be introduced later as well. 

## Basic manipulations

In the following, I will introduce you to a number of different operations that can be performed on strings.

### Changing the case of the words

A basic operation is changing words' cases.

```{r}
library(tidyverse) #stringr is part of the core tidyverse
library(stringr)
str_to_lower(vector_of_strings)
str_to_upper(vector_of_strings)
str_to_title(vector_of_strings)
str_to_sentence(vector_of_strings)
```

### Determining a string's length

Determining the string's number of characters goes as follows: 

```{r}
str_length(vector_of_strings)
```

### Extracting particular characters

Characters can be extracted (by position) using `str_sub`

```{r}
str_sub(vector_of_strings, start = 1, end = 5) # extracting first to fifth character
str_sub(vector_of_strings, start = -5, end = -1) # extracting fifth-to-last to last character
```

You can also use `str_sub()` to replace strings. E.g., to replace the last character by a full stop, you can do the following:

```{r}
str_sub(vector_of_strings, start = -1) <- "."
vector_of_strings
```

However, in everyday use, you would probably go with `str_replace()` and regular expressions.

### Concatenating strings

Similar to how `c()` puts together different elements (or vectors of length 1) and other vectors into a single vector, `str_c()` can be used to concatenate several strings into a single string. This can, for instance, be used to write some birthday invitations.

```{r}
names <- c("Inger", "Peter", "Kalle", "Ingrid")

str_c("Hi", names, "I hope you're doing well. As per this letter, I invite you to my birthday party.")
```

Well, this looks kind of ugly, as there are no spaces, and commas are lacking as well. You can fix that by determining a separator using the `sep` argument. 

```{r}
str_c("Hi", names, "I hope you're doing well. As per this letter, I invite you to my birthday party.", sep = ", ")
```

You could also collapse the strings contained in a vector together into one single string using the `collapse` argument. 

```{r}
str_c(names, collapse = ", ")
```

This can also be achieved using the `str_flatten()` function.

```{r}
str_flatten(names, collapse = ", ")
```

### Repetition

Repeating (or duplicating) strings is performed using `str_dup()`. The function takes two arguments: the string to be duplicated and the number of times.

```{r}
str_dup("felix", 2)
str_dup("felix", 1:3)
str_dup(names, 2)
str_dup(names, 1:4)
```

### Removing unnecessary whitespaces

Often text contains unnecessary whitespaces. 
```{r}
unnecessary_whitespaces <- c("    on the left", "on the right    ", "    on both sides   ", "   literally    everywhere  ")
```

Removing the ones at the beginning of the end of a string can be accomplished using `str_trim()`.

```{r}
str_trim(unnecessary_whitespaces, side = "left")
str_trim(unnecessary_whitespaces, side = "right")
str_trim(unnecessary_whitespaces, side = "both") # the default option
```

`str_trim()` could not fix the last string though, where unnecessary whitespaces were also present in between words. Here, `str_squish` is more appropriate. It removes leading or trailing whitespaces as well as duplicated ones in between words.

```{r}
str_squish(unnecessary_whitespaces)
```

## Regular expressions

```{r echo=FALSE}
vembedr::embed_youtube("EAxIcgbGWaw")
```

Up to now, you have been introduced to the more basic functions of the `stringr` package. Those are useful, for sure, yet limited. However, to make use of the full potential of `stringr`, you will first have to acquaint yourself with regular expressions (also often abbreviated as "regex" with plural "regexes"). 

Those regular expressions are patterns that can be used to describe certain strings. Exemplary use cases of regexes are the identification of phone numbers, email addresses, or whether a password you choose on a web page consists of enough characters, an uppercase character, and at least one special character. Hence, if you want to replace certain words with another one, you can write the proper regex and it will identify the strings you want to replace, and the `stringr` functions (i.e., `str_replace()`) will take care of the rest.

Before you dive into regexes, beware that they are quite complicated at the beginning (honestly, I was quite overwhelmed when I encountered them first). Yet, mastering them is very rewarding and will definitely pay off in the future.

### Literal characters

The most basic regex patterns consist of literal characters only. `str_view()` tells you which parts of a string match a pattern is present in the element. 

```{r}
five_largest_cities <- c("Stockholm", "Göteborg", "Malmö", "Uppsala", "Västerås")
```

```{r eval=FALSE, include=FALSE}
str_c(five_largest_cities, collapse = "\n") %>% cat()
```

Note that regexes are case-sensitive.

```{r}
str_view(five_largest_cities, "stockholm")
```

```{r}
str_view(five_largest_cities, "Stockholm")
```

They also match parts of words:

```{r}
str_view(five_largest_cities, "borg")
```

Moreover, they are "greedy," they only match the first occurrence (in "Stockholm"):

```{r}
str_view(five_largest_cities, "o")
```

This can be addressed in the `stringr` package by using `str_._all()` functions -- but more on that later.

If you want to match multiple literal characters (or words, for that sake), you can connect them using the `|` meta character (more on meta characters later).

```{r}
str_view(five_largest_cities, "Stockholm|Göteborg")
```

Every letter of the English alphabet (or number/or combination of those) can serve as a literal character. Those literal characters *match themselves*. This is, however, not the case with the other sort of characters, so-called meta characters.

### Metacharacters

When using regexes, the following characters are considered meta characters and have a special meaning:

`. \ | ( ) { } [ ] ^ $ - * + ?`

#### The wildcard

Did you notice how I used the dot to refer to the entirety of the `str_._all()` functions? This is basically what the `.` meta-character does: it matches every character except for a new line. The first call extracts all function names from the `stringr` package, the second one shows the matches (i.e., the elements of the vector where it can find the pattern).

```{r}
stringr_functions <- ls("package:stringr")

str_detect(stringr_functions, "str_._all")
```

```{r eval=FALSE, include=FALSE}
str_c(stringr_functions, collapse = "\n") %>% cat()
```

Well, as you can see, there are none. This is due to the fact that the `.` can only replace one character. We need some sort of multiplier to find them. The ones available are:

* `?` -- zero or one
* `*` -- zero or more
* `+` -- one or more
* `{n}` -- exactly n
* `{n,}` -- n or more
* `{n,m}` -- between n and m

In our case, the appropriate one is `+`:

```{r}
str_detect(stringr_functions, "str_.+_all")
```

However, if you want to match the character dot? This problem may arise when searching for clock time. A naive regex might look like this:

```{r}
vectors_with_time <- c("13500", "13M00", "13.00")

str_detect(vectors_with_time, "13.00")
```

```{r eval=FALSE, include=FALSE}
str_c(vectors_with_time, collapse = "\n") %>% cat()
```

Yet, it matches everything. We need some sort of literal dot. Here, the metacharacter `\` comes in handy. By putting it in front of the metacharacter, it no longer has its special meaning and is interpreted as a literal character. This procedure is referred to as "escaping." Hence, `\` is also referred to as the "escape character." Note that you will need to escape `\` as well, and therefore it will look like this: `\\.`. 

```{r}
str_detect(vectors_with_time, "13\\.00")
```

### Sets of characters

You can also define sets of multiple characters using the `[ ]` meta characters. This can be used to define multiple possible characters that can appear in the same place. 

```{r}
sp_ce <- c("spice", "space")

str_view(sp_ce, "sp[ai]ce")
```

```{r eval=FALSE, include=FALSE}
str_c(sp_ce, collapse = "\n") %>% cat()
```

You can also define certain ranges of characters using the `-` metacharacter:

Same holds for numbers:

```{r}
american_phone_number <- "(555) 555-1234"

str_view(american_phone_number, "\\([:digit:]{3}\\) [0-9]{3}-[0-9]{4}")
```

There are also predefined sets of characters, for instance, digits or letters, which are called *character classes*. You can find them on the [`stringr` cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf). 

Furthermore, you can put almost every meta character inside the square brackets without escaping them. This does not apply to the caret (`^`) in the first position, the dash `-`, the closing square bracket `]`, and the backslash `\`.

```{r}
str_view(vector_of_strings, "[.]")
```

```{r eval=FALSE, include=FALSE}
str_c(vector_of_strings, collapse = "\n") %>% cat()
```

#### Negating sets of characters

Sometimes you will also want to exclude certain sets of characters or words. In order to achieve this, you can use the `^` meta character at the beginning of the range or set you are defining. 

```{r}
str_view(sp_ce, "sp[^i]ce")
```

### Anchors

There is also a way to define whether you want the pattern to be present in the beginning `^` or at the end `$` of a string. `sentences` are a couple of (i.e., 720) predefined example sentences. If I were now interested in the number of sentences that begin with a "the," I could write the following regex:

```{r}
shortened_sentences <- sentences[1:10]

str_view(shortened_sentences, "^The") 
```

```{r eval=FALSE, include=FALSE}
str_c(shortened_sentences, collapse = "\n") %>% cat()
```

If I wanted to know how many start with a "The" and end with a full stop, I could do this one:

```{r}
str_view(shortened_sentences, "^The.+\\.$") 
```

#### Boundaries

Note that right now, the regex also matches the sentence which starts with a "These." In order to address this, I need to tell the machine that it should only accept a "The" if there starts a new word thereafter. In regex syntax, this is done using so-called boundaries. Those are defined as `\b` as a word boundary and `\B` as no word boundary. (Note that you will need an additional escape character as you will have to escape the escape character itself.)

In my example, I would include the former if I were to search for sentences that begin with a single "The" and the latter if I were to search for sentences that begin with a word that starts with a "The" but are not "The" -- such as "These."

```{r}
str_view(shortened_sentences, "^The\\b.+\\.$") 
str_view(shortened_sentences, "^The\\B.+\\.$") 
```

#### Lookarounds

A final common task is to extract certain words or values based on what comes before or after them. Look at the following example:

```{r}
heights <- c("1m30cm", "2m01cm", "3m10cm")
```

```{r eval=FALSE, include=FALSE}
str_c(heights, collapse = "\n") %>% cat()
```

Here, in order to identify the height in meters, the first task is to identify all the numbers that are followed by an "m". The regex syntax for this looks like this: `A(?=pattern)` with `A` being the entity that is supposed to be found (hence, in this case, [0-9]+).

```{r}
str_view(heights, "[0-9]+(?=m)")
```

The second step now is to identify the centimeters. This could of course be achieved using the same regex and replacing `m` with `cm`. However, we can also harness a so-called negative look ahead `A(?!pattern)`, a so-called look behind `(?<=pattern)A`. The negative counterpart, the negative look behind `(?<!pattern)A` could be used to extract the meters.

The negative lookahead basically returns everything that is not followed by the defined pattern. The look behind returns everything that is preceded by the pattern, the negative look behind returns everything that is not preceded by the pattern.

In the following, I demonstrate how you could extract the centimeters using negative look ahead and look behind.

```{r}
str_view(heights, "[0-9]+(?!m)") # negative look ahead
```

```{r}
str_view(heights, "(?<=m)[0-9]+") # look behind
```

## More advanced string manipulation

Now that you have learned about regexes, you can unleash the full power of `stringr`.

The basic syntax of a `stringr` function looks as follows: `str_.*(string, regex(""))`. Some `stringr` functions also have the suffix `_all` which implies that they perform the operation not only on the first match ("greedy") but on every match.

In order to demonstrate the different functions, I will again rely on the subset of example sentences.

### Detect matches

`str_detect` can be used to determine whether a certain pattern is present in the string.

```{r}
str_detect(shortened_sentences, "The\\b")
```

This also works very well in a `dplyr::filter()` call. Finding all action movies in the IMDB data set can be solved like this:

```{r}
imdb_raw <- read_csv("data/imdb2006-2016.csv")
imdb_raw %>% 
  filter(str_detect(Genre, "Action"))
```

If you want to know whether there are multiple matches present in each string, you can use `str_count`. Here, it might by advisable to set the `ignore_case` option to `TRUE`:

```{r}
str_count(shortened_sentences, regex("the\\b", ignore_case = TRUE))
```

If you want to locate the match in the string, use `str_locate`. This returns a matrix, which is basically a vector of multiple dimensions.

```{r}
str_locate(shortened_sentences, regex("The\\b", ignore_case = TRUE))
```

Moreover, this is a good example for the greediness of `stringr` functions. Hence, it is advisable to use `str_locate_all` which returns a list with one matrix for each element of the original vector:

```{r}
str_locate_all(shortened_sentences, regex("The\\b", ignore_case = TRUE))
```

### Mutating strings

Mutating strings usually implies the replacement of certain elements (e.g., words) with other elements (or removing them, which is basically a special case of replacing them). In `stringr` this is performed using `str_replace(string, pattern, replacement)` and `str_replace_all(string, pattern, replacement)`. 

If I wanted, for instance, to replace the first occurrence of "m" letters with "meters," I would go about this the following way:

```{r}
str_replace(heights, "m", "meters")
```

Note that `str_replace_all` would have lead to the following outcome:

```{r}
str_replace_all(heights, "m", "meters")
```

However, I also want to replace the "cm" with "centimeters," hence, I can harness another feature of `str_replace_all()`:

```{r}
str_replace_all(heights, c("m" = "meters", "cm" = "centimeters"))
```

What becomes obvious is that a "simple" regex containing just literal characters more often than not does not suffice. It will be your task to fix this. And while on it, you can also address the meter/meters problem -- a "1" needs meter instead of meters. Another feature is that the replacements are performed in order. You can harness this for solving the problem.

<details>
  <summary>Solution. Click to expand!</summary>
Solution:
```{r}
str_replace_all(heights, c("(?<=[2-9]{1})m" = "meters", "(?<=[0-9]{2})m" = "meters", "(?<=1)m" = "meter", "(?<=01)cm$" = "centimeter", "cm$" = "centimeters"))
```
</details>

### Extracting text

`str_extract(_all)()` can be used to extract matching strings. In the `mtcars` data set, the first word describes the car brand. Here, I harness another regexp, the `\\w` which stands for any word character. Its opponent is `\\W` for any non-word character.

```{r}
mtcars %>% 
  rownames_to_column(var = "car_model") %>% 
  transmute(manufacturer = str_extract(car_model, "^\\w+\\b"))
```

### Split vectors

Another use case here would have been to split it into two columns: manufacturer and model. One approach would be to use `str_split()`. This function splits the string at every occurrence of the predefined pattern. In this example, I use a word boundary as the pattern:

```{r}
manufacturer_model <- rownames(mtcars)
str_split(manufacturer_model, "\\b") %>% 
  head()
```

This outputs a list containing the different singular words/special characters. This doesn't make sense in this case. Here, however, the structure of the string is always roughly the same: "\\[manufacturer\\]\\[ \\]\\[model description\\]". Moreover, the manufacturer is only one word. Hence, the task can be fixed by splitting the string after the first word, which should indicate the manufacturer. This can be accomplished using `str_split_fixed()`. Fixed means that the number of splits is predefined. This returns a matrix that can easily become a tibble.

```{r}
str_split_fixed(manufacturer_model, "(?<=\\w)\\b", n = 2) %>% 
  as_tibble() %>% 
  rename(manufacturer = V1,
         model = V2) %>% 
  mutate(model = str_squish(model))
```

## Further links

* The [`stringr` cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf).
* A [YouTube video](https://www.youtube.com/watch?v=NvHjYOilOf8) on regexes by Johns Hopkins professor Roger Peng.
* And a [chapter](https://bookdown.org/rdpeng/rprogdatascience/regular-expressions.html#the-stringr-package) by Roger Peng.
* A [website for practicing regexes](https://regexone.com).
* You can also consult the `introverse` package if you need help with the packages covered here -- `introverse::show_topics("stringr")` will give you an overview of the `stringr` package's functions, and `get_help("name of function")` will help you with the respective function.

## Exercises

1. Find all Mercedes in the `mtcars` data set.
2. Write regexes that correctly determine the country of origin of different phone numbers. 
  a. By country code -- +46 for Sweden, +49 for Germany, +33 for France. (Examples: "+46 71-738 25 33" (Swedish), "+33 6 51 95 24 27" (France -- note that mobile numbers in France always start with 6 or 7), "+49 173 7462526"
  b. When the country code is missing, by format. (Examples: "0 71-738 25 33" (Swedish -- always start with 7 after leading 0 or country code), "0 6 51 95 24 27" (France -- always start with 6 or 7 after leading 0 or country code), "0173 7462526" (Germany -- always start with 1 after leading 0 or country code). Test yourself by determining the country of origin of the following numbers: 0178 5693445, 0 76-426 49 14, 0160 90609495, 0 7 43 34 23 23. 
  c. Replace the leading zeroes in b. with the correct country codes. 
3. Remember the vector of heights? 
    a. How can you extract the meters using the negative look behind?
    b. Bring it into numeric format (i.e., `your_solution == c(1.3, 2.01, 3.1)`) using regexes and `stringr` commands -- and, subsequently, `as.numeric()`.
4. For the advanced folks: Write a function that determines if a password is safe. It's supposed to return "high safety" if the password contains at least one uppercase letter, lowercase letter(s), and a  number, "medium safety" if two of the former criteria are met, and "low safety" if it contains only lowercase letters. Passwords need to be of length 8, otherwise, the function needs to state "unacceptable".
```{r}
test_numbers_nocountrycode <- c("0178 5693445", "0 76-426 49 14", "0160 90609495", "0 7 43 34 23 23")
```

```{r eval=FALSE}
high_safety <- "Password1234"
medium_safety <- "password1234"
low_safety <- "password"
not_accepted <- "1jj2345"
```

5. Take the [IMDb file](https://www.dropbox.com/s/81o3zzdkw737vt0/imdb2006-2016.csv?dl=0) and split the `Genre` column into different columns (hint: look at the `tidyr::separate()` function). How would you do it if `Genre` were a vector using `str_split_fixed()`?

<details>
  <summary>Solution. Click to expand!</summary>

```{r eval=FALSE}
#1
mtcars %>% 
  rownames_to_column("model") %>% 
  filter(str_detect(model, "Merc"))

#2
#a
phone_numbers_countrycode <- c("+46 71-738 25 33", "+33 6 51 95 24 27", "+49 173 7462526")
swedish <- phone_numbers_countrycode[str_detect(phone_numbers_countrycode, "^\\+46")]
german <- phone_numbers_countrycode[str_detect(phone_numbers_countrycode, "^\\+49")]
french <- phone_numbers_countrycode[str_detect(phone_numbers_countrycode, "^\\+33")]

#b
phone_numbers_no_countrycode <- c("0178 5693445", "0 76-426 49 14", "0160 90609495", "0 7 43 34 23 23")
swedish <- phone_numbers_no_countrycode[str_detect(phone_numbers_no_countrycode, "0 7[0-9]\\-[0-9]{3} [0-9]{2} [0-9]{2}")]
german <- phone_numbers_no_countrycode[str_detect(phone_numbers_no_countrycode, "01[0-9]{2,3} [0-9]{6}")]
french <- phone_numbers_no_countrycode[str_detect(phone_numbers_no_countrycode, "0 [67] [0-9]{2} [0-9]{2} [0-9]{2} [0-9]{2}")]

#c 
replace_countrycode <- function(phone_number){
  if(str_detect(phone_number, "0 7[0-9]\\-[0-9]{3} [0-9]{2} [0-9]{2}")) return(str_replace(phone_number, "^0", "+46"))
  if(str_detect(phone_number, "01[0-9]{2,3} [0-9]{6}")) return(str_replace(phone_number, "^0", "+49 "))
  if(str_detect(phone_number, "0 [67] [0-9]{2} [0-9]{2} [0-9]{2} [0-9]{2}")) return(str_replace(phone_number, "^0", "+33"))
}

#check it
map_chr(phone_numbers_no_countrycode, replace_countrycode)

#3
heights <- c("1m30cm", "2m01cm", "3m10cm")

#a
meters <- str_extract(heights, "(?<!m)[0-9]")

#b
for_test <- heights %>% 
  str_replace("(?<=[0-9])m", "\\.") %>% 
  str_replace("cm", "") %>% 
  as.numeric() 

for_test == c(1.3, 2.01, 3.1)

#4
high_safety <- "Password1234"
medium_safety <- "password1234"
low_safety <- "password"
not_accepted <- "12345"

return_safety <- function(password){
  if(str_length(password) < 8) return("unacceptable")
  if(str_detect(password, "[0-9]") & 
     str_detect(password, "[A-Z]") & 
     str_detect(password, "[a-z]")) {
    return("high safety")
  }
  if(str_detect(password, "[0-9]") & str_detect(password, "[a-z]") |
     str_detect(password, "[A-Z]") & str_detect(password, "[a-z]")) {
    return("medium safety")
  }
  if(str_detect(password, "[a-z]")) return("low safety")
}

#return_safety(high_safety)
#return_safety(medium_safety)
#return_safety(low_safety)
#return_safety(unacceptable)

#5
imdb <- read_csv("imdb2006-2016.csv")

imdb %>% 
  separate(Genre, sep = ",", into = c("genre_1", "genre_2", "genre_3"))

imdb$Genre %>% 
  str_split_fixed(pattern = ",", 3)
```

</details>
